---
title: "Clustering Analysis of Disease Concern Levels"
author: "Andrea Pedot"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: true
---

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(ggfortify)
library(plotly)
library(ggplot2)
library(ggdendro)
library(cluster)
library(factoextra)
library(targets)

# Load custom pipeline data and functions
tar_source()
survey <- tar_read(survey)
```

## Introduction

In this analysis, we explore patterns in public concern levels about different diseases using clustering techniques. Specifically, we use:

-   **Principal Component Analysis (PCA)** for dimensionality reduction and visualization.

-   **K-Means Clustering** to group diseases with similar concern profiles.

-   **Hierarchical Clustering** to examine disease similarity through dendrograms.

## Data Preparation

We start by transforming the `survey` data into long format and filtering for only the `"Very concerned"` responses.

```{r}
long <- pivot_disease_data(survey)

wide <- long |>
  dplyr::filter(response == "Very concerned") |>  # Optional: filter by response
  mutate(question_response = paste(question, response, sep = " - ")) |>
  select(disease, question_response, prop) |>
  pivot_wider(names_from = question_response, values_from = prop, values_fill = 0)

summary(wide)

```

## Principal Component Analysis (PCA)

We apply PCA to reduce the dimensionality of the data for easier visualization.

```{r}
pca_input <- wide |>
  select(-disease) |>
  as.matrix()

pca_result <- prcomp(pca_input, scale. = TRUE)
```

To get a better sense of spatial relationships, we visualize the first three principal components using **Plotly** in 3D.

```{r}
pca_df <- as_tibble(pca_result$x[, 1:3]) |>
  mutate(disease = wide$disease)

plot_ly(pca_df, x = ~PC1, y = ~PC2, z = ~PC3,
        type = 'scatter3d', mode = 'text+markers',
        text = ~disease,
        marker = list(size = 4)) |>
  layout(title = "PCA of Disease Similarity (3D)")
```

## K-Means Clustering

We use K-means clustering to identify groups of diseases with similar concern profiles.

```{r}
# Recreate matrix with disease as row names
mat <- wide |>
  column_to_rownames("disease") |>
  as.matrix()

# Scale the data
scaled_mat <- scale(mat)

set.seed(123)
# Perform K-means with k = 2
kmeans_result <- kmeans(scaled_mat, centers = 2)
fviz_cluster(kmeans_result, data = scaled_mat,
             geom = "text", labelsize = 12,
             repel = TRUE,
             main = "K-Means Clustering of Diseases (k = 2)",
             ggtheme = theme_minimal())

# Perform K-means with k = 4
kmeans_result <- kmeans(scaled_mat, centers = 4)
fviz_cluster(kmeans_result, data = scaled_mat,
             geom = "text", labelsize = 12,
             repel = TRUE,
             main = "K-Means Clustering of Diseases (k = 4)",
             ggtheme = theme_minimal())

# Perform K-means with k = 5
kmeans_result <- kmeans(scaled_mat, centers = 5)
fviz_cluster(kmeans_result, data = scaled_mat,
             geom = "text", labelsize = 12,
             repel = TRUE,
             main = "K-Means Clustering of Diseases (k = 5)",
             ggtheme = theme_minimal())

# Perform K-means with k = 6
kmeans_result <- kmeans(scaled_mat, centers = 6)
fviz_cluster(kmeans_result, data = scaled_mat,
             geom = "text", labelsize = 12,
             repel = TRUE,
             main = "K-Means Clustering of Diseases (k = 6)",
             ggtheme = theme_minimal())
```

**Section: Determine Optimal Number of Clusters**

```{r}
# Elbow Method 
fviz_nbclust(scaled_mat, kmeans, method = "wss") +
  ggtitle("Elbow Method for Optimal k") 
# Silhouette Method 
fviz_nbclust(scaled_mat, kmeans, method = "silhouette") +
  ggtitle("Silhouette Method for Optimal k")
```

These plots help determine the best number of clusters:

-   The **elbow** in the WSS (Within Sum of Squares) plot suggests a good `k`.

-   The **silhouette** method shows how well-separated the clusters are.

## Hierarchical Clustering

We also explore disease relationships using hierarchical clustering and a dendrogram.

```{r}
# Compute distance matrix
dist_mat <- dist(mat)

# Perform hierarchical clustering using Ward's method
hc <- hclust(dist_mat, method = "ward.D2")

# Plot dendrogram
ggdendrogram(hc, rotate = FALSE, size = 2) +
  ggtitle("Hierarchical Clustering of Diseases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Conclusion

This analysis shows how clustering techniques can be used to uncover patterns in public concern about diseases. PCA helped reduce dimensionality for visualization, while both K-means and hierarchical clustering offered complementary views on disease similarity.
